Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-02-13 WholePlotTrajectories.Rmd
intro

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, and ensure cultural and human well-being.
They regulate as well the local and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current global changes increase the pressure on remaining natural forests [@Morales-Hidalgo2015].
These threats affect the natural disturbance regime that shape and maintain the structure, composition and functioning of tree communities [@Schnitzer2001;@Anderson-Teixeira2013;@Sist2015].
It is now urgent to understand the processes underlying community response to disturbance, and thus to be able to anticipate the fate of tropical forests in the current context.
In tropical forests, disturbance maintains a forest environment and initiates a suit of changes in the structure and functioning of communities consituting a secondary succession.
Disturbance change both community abiotic and biotic environments, through modifications in the fluxes of light, heat and water [@Goulamoussene2017]. 
For now, post-disturbance trajectories has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Models based on observed trajectories then assessed community time to recovery and identified some of the determinants of post-disturbance trajectories, like forest structure, forest composition, environmental parameters and disturbance characteristics [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories have not been as thoroughly understood [@Guitet2018;@Molino2001].
The variety of species response to disturbance entail different processes like colonisation, facilitation, coexistence, etc, which succession shapes biodiversity post-disturbance trajectories [@Lindenmayer2012;@Garcia_florez2017].
The rate, endpoint and duration of the trajectories depend on the intensity and frequency of disturbance: a current objective of tropical ecology is to highlight the link betwenn biodiversity and disturbance and anticipate community trajectories [@Chazdon2003a;@Willig2018].

An early conceptual basis of the linkage between biodiversity and disturbance is the Intermediate Disturbance Hypothesis (IDH).
The IDH assumes a relationship between community diversity and the intensity and frequency of disturbance events [@Connell1978].
The theory and states that community diversity peaks at intermediate level of disturbance.
The fluctuations of environmental conditions after disturbance are assumed to prevent competitive exclusion in fostering both competitively superior species and fast colonizers [@Shea2004;@Pulsford2016].
In highly diverse tropical forests, however, observations of the IDH are often blurry compared to theoretical expectations [@Hugues2007;@Sheil2003;@Norden2017], and the validation of the theory remains controversial [@Hubbell2001;@Fox2013;@Sheil2013].
As previously highlighted, there is a variety of ecological processes emerging after disturbance, and the pattern of the IDH (the diversity peak under intermediate disturbance) can result from several such processes.
Disentangle the processes involved and their succession in time would allow discussing the conditions for the application of the IDH, its scope in time, and its implication for the fate of forests [@Sheil2003;@Shea2004].

Fully analyse post-disturbance trajectories requires a set of metrics, to grasp all aspects of community changes [@Sheil2003;@Shea2004;@Mayfield2010].
The analysis should first consider community composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006] and assesses the different species fostered by the disturbance.
Diversity metrics encompassing taxonomic richness and evenness would assess the changes community abundance distribution. 
A functional approach accounting for species biological attributes and ecological strategy, would directly link community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach also encompasses the analysis of functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, (i) from community trajectories we examined their taxonomic and functional resilience, (ii) discussed the scope of the IDH and the underlying ecological processes, and (iii) analyzed community time to recovery.

@@

## Decoupled taxonomic and functional trajectories

@@

## A new perspective on the intermediate disturbance hypothesis

Community taxonomic richness and evenness trajectories were determined by the disturbance intensity, ranging from a limited and temporary impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold for which the taxonomic richness was maximized and the taxonomic evenness was resilient. 
The disturbance intensity determined the balance in the community between pre-disturbance surviving trees, and trees recruited afterward. 
Below a 20%-25% AGB loss, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community [@Bongers2009].
The recruitment of pionners, infrequent or absent before disturbance, then increased the taxonomic richness all the more so that the disturbance was intense [@Martin2015; @Chaudhary2016].
As these pioneers became more dominant they balanced the usual hyper-dominance of tropical forests, hence temporarily increasing the taxonomic evenness [@Baraloto2012a].
Above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
The taxonomic richness then decreased following intense disturbance, all the more so that the disturbance intensity was high [@Molino2001]. 
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity besides highlighted an intermediate time after disturbance for which the taxonomic evenness was maximized.
As already observed in the Guiana shield [@Baraloto2012a] and in Borenan tropical forests [@Cannon1998],
taxonomic evenness followed hump-shaped trajectories with a peak when the recruitment was balanced between pioneers and late-successional species. 
This time then corresponded to the match between the generation time of pionneers, first to dominate the post-disturbance recruitment, and the time for competitive exclusion to emerge, which leads to the dominance of late-successional species.

>>>>

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, and ensure cultural and human well-being.
They regulate as well the local and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current global changes increase the pressure on remaining natural forests [@Morales-Hidalgo2015].
These threats affect the natural disturbance regime that shape and maintain the structure, composition and functioning of tree communities [@Schnitzer2001;@Anderson-Teixeira2013;@Sist2015].
It is now urgent to understand the processes underlying community response to disturbance, and thus to be able to anticipate the fate of tropical forests in the current context.
In tropical forests, disturbance maintains a forest environment and initiates a suit of changes in the structure and functioning of communities consituting a secondary succession.
Disturbance change both community abiotic and biotic environments, through modifications in the fluxes of light, heat and water [@Goulamoussene2017]. 
For now, post-disturbance trajectories has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Models based on observed trajectories then assessed community time to recovery and identified some of the determinants of post-disturbance trajectories, like forest structure, forest composition, environmental parameters and disturbance characteristics [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories have not been as thoroughly understood [@Guitet2018;@Molino2001].
Given the variety of species response to disturbance and the huge diversity of tropical forests, biodiversity trajectories following disturbance might be very diverse [@Lindenmayer2012;@Garcia_florez2017].
We assumed that the rate, endpoint and duration of the trajectories depend on the intensity and frequency of disturbance: we then aimed to highlight the link between biodiversity and disturbance and thus anticipate community post-disturbance trajectories [@Chazdon2003a;@Willig2018].

An early conceptual basis of the linkage between biodiversity and disturbance is the Intermediate Disturbance Hypothesis (IDH).
The IDH assumes a relationship between community diversity and the intensity and frequency of disturbance events [@Connell1978].
The theory and states that community diversity peaks at intermediate level of disturbance.
The fluctuations of environmental conditions after disturbance are assumed to prevent competitive exclusion in fostering both competitively superior species and fast colonizers [@Shea2004;@Pulsford2016].
In tropical forests, however, observations of the IDH are often blurry compared to theoretical expectations [@Hugues2007;@Sheil2003;@Norden2017], and the validation of the theory remains controversial [@Hubbell2001;@Fox2013;@Sheil2013].
The IDH may result from several processes and in the highly diverse tree communities its pattern might differ from those recognized otherwise.
Most of the studies rely on a single diversity metric, often the taxonomic richness [@Molino2001], which is limited to grasp all facets of the diversity-disturbance relationship.
In this study we aimed as well to clarify the scope of the IDH and discuss its implication for the fate of tropical forests [@Sheil2003;@Shea2004].

The analyse of community response to disturbance requires a set of metrics, to grasp all aspects of community changes [@Sheil2003;@Shea2004;@Mayfield2010].
The analysis should first consider community composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006] and assesses the different species fostered by the disturbance.
Diversity metrics encompassing taxonomic richness and evenness would assess the changes community abundance distribution. 
A functional approach accounting for species biological attributes and ecological strategy, would directly link community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach also encompasses the analysis of functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, (i) from post-disturbance trajectories we examined community taxonomic and functional response, (ii) then we discussed the scope of the IDH in taxonomic and functional facets and in time after disturbance, and (iii) we analyzed community resilience and time to recovery.

@@

## Decoupled taxonomic and functional composition trajectories

@@

## The scope of the intermediate disturbance hypothesis

Community taxonomic richness and evenness trajectories were determined by the disturbance intensity, ranging from a limited and temporary impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold for which the taxonomic richness was maximized and the taxonomic evenness remained resilient. 
The disturbance intensity determined the balance in the community between pre-disturbance surviving trees, and trees recruited afterward. 
Below a 20%-25% AGB loss, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community [@Bongers2009].
The recruitment of pionners, infrequent or absent before disturbance, then increased the taxonomic richness all the more so that the disturbance was intense [@Martin2015; @Chaudhary2016].
As these pioneers became more dominant they balanced the usual hyper-dominance of tropical forests, hence temporarily increasing the taxonomic evenness [@Baraloto2012a].
Above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
The taxonomic richness then decreased following intense disturbance, all the more so that the disturbance intensity was high [@Molino2001]. 
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity besides highlighted an intermediate time after disturbance, around 15-20 years, for which the taxonomic evenness was maximized.
As already observed in the Guiana shield [@Baraloto2012a] and in Borenan tropical forests [@Cannon1998],
taxonomic evenness followed hump-shaped trajectories with a peak when the recruitment was balanced between pioneers and late-successional species. 
This intermediate time corresponded to the match between the generation time of pionneers, first to dominate the post-disturbance recruitment, and the time for competitive exclusion to emerge, which leads to the dominance of late-successional species.

### 2019-02-13 WholePlotTrajectories.tex
intro

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. These threats affect the natural
disturbance regime that shape and maintain the structure, composition
and functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. It is now urgent
to understand the processes underlying community response to
disturbance, and thus to be able to anticipate the fate of tropical
forests in the current context. In tropical forests, disturbance
maintains a forest environment and initiates a suit of changes in the
structure and functioning of communities consituting a secondary
succession. Disturbance change both community abiotic and biotic
environments, through modifications in the fluxes of light, heat and
water \citep{Goulamoussene2017}. For now, post-disturbance trajectories
has been largely studied through forest structural parameters such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Models based on observed
trajectories then assessed community time to recovery and identified
some of the determinants of post-disturbance trajectories, like forest
structure, forest composition, environmental parameters and disturbance
characteristics \citep{Herault2018}. Regarding community diversity and
composition, however, post-disturbance trajectories have not been as
thoroughly understood \citep{Guitet2018, Molino2001}. The variety of
species response to disturbance entail different processes like
colonisation, facilitation, coexistence, etc, which succession shapes
biodiversity post-disturbance trajectories
\citep{Lindenmayer2012, Garcia_florez2017}. The rate, endpoint and
duration of the trajectories depend on the intensity and frequency of
disturbance: a current objective of tropical ecology is to highlight the
link betwenn biodiversity and disturbance and anticipate community
trajectories \citep{Chazdon2003a, Willig2018}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events \citep{Connell1978}. The theory and
states that community diversity peaks at intermediate level of
disturbance. The fluctuations of environmental conditions after
disturbance are assumed to prevent competitive exclusion in fostering
both competitively superior species and fast colonizers
\citep{Shea2004, Pulsford2016}. In highly diverse tropical forests,
however, observations of the IDH are often blurry compared to
theoretical expectations \citep{Hugues2007, Sheil2003, Norden2017}, and
the validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. As previously highlighted,
there is a variety of ecological processes emerging after disturbance,
and the pattern of the IDH (the diversity peak under intermediate
disturbance) can result from several such processes. Disentangle the
processes involved and their succession in time would allow discussing
the conditions for the application of the IDH, its scope in time, and
its implication for the fate of forests \citep{Sheil2003, Shea2004}.

Fully analyse post-disturbance trajectories requires a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
\citep{Lavorel2002, Bellwood2006} and assesses the different species
fostered by the disturbance. Diversity metrics encompassing taxonomic
richness and evenness would assess the changes community abundance
distribution. A functional approach accounting for species biological
attributes and ecological strategy, would directly link community
diversity, composition and redundancy to ecosystem functioning and
environmental constraints \citep{Violle2007b, Baraloto2012a}. In that
respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, (i) from community trajectories we examined their
taxonomic and functional resilience, (ii) discussed the scope of the IDH
and the underlying ecological processes, and (iii) analyzed community
time to recovery.

@@

\subsection{Decoupled taxonomic and functional
trajectories}\label{decoupled-taxonomic-and-functional-trajectories}

@@

\subsection{A new perspective on the intermediate disturbance
hypothesis}\label{a-new-perspective-on-the-intermediate-disturbance-hypothesis}

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
for which the taxonomic richness was maximized and the taxonomic
evenness was resilient. The disturbance intensity determined the balance
in the community between pre-disturbance surviving trees, and trees
recruited afterward. Below a 20\%-25\% AGB loss, the trees surviving
after disturbance remained numerous enough to maintain the high
taxonomic richness of the pre-disturbance community \citep{Bongers2009}.
The recruitment of pionners, infrequent or absent before disturbance,
then increased the taxonomic richness all the more so that the
disturbance was intense \citep{Martin2015, Chaudhary2016}. As these
pioneers became more dominant they balanced the usual hyper-dominance of
tropical forests, hence temporarily increasing the taxonomic evenness
\citep{Baraloto2012a}. Above the intensity threshold, the taxonomic
richness of surviving trees was too low to be offset by the recruitment
of pioneers. In the Guiana shield indeed, the pool of true pioneers
specifically recruited after disturbance is restricted to a few common
genera (e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)
\citep{Guitet2018}. The taxonomic richness then decreased following
intense disturbance, all the more so that the disturbance intensity was
high \citep{Molino2001}. At the same time, pioneers became persistently
dominant and prevented a return towards the initial evenness and the
recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity
besides highlighted an intermediate time after disturbance for which the
taxonomic evenness was maximized. As already observed in the Guiana
shield \citep{Baraloto2012a} and in Borenan tropical forests
\citep{Cannon1998}, taxonomic evenness followed hump-shaped trajectories
with a peak when the recruitment was balanced between pioneers and
late-successional species. This time then corresponded to the match
between the generation time of pionneers, first to dominate the
post-disturbance recruitment, and the time for competitive exclusion to
emerge, which leads to the dominance of late-successional species.

>>>>

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. These threats affect the natural
disturbance regime that shape and maintain the structure, composition
and functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. It is now urgent
to understand the processes underlying community response to
disturbance, and thus to be able to anticipate the fate of tropical
forests in the current context. In tropical forests, disturbance
maintains a forest environment and initiates a suit of changes in the
structure and functioning of communities consituting a secondary
succession. Disturbance change both community abiotic and biotic
environments, through modifications in the fluxes of light, heat and
water \citep{Goulamoussene2017}. For now, post-disturbance trajectories
has been largely studied through forest structural parameters such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Models based on observed
trajectories then assessed community time to recovery and identified
some of the determinants of post-disturbance trajectories, like forest
structure, forest composition, environmental parameters and disturbance
characteristics \citep{Herault2018}. Regarding community diversity and
composition, however, post-disturbance trajectories have not been as
thoroughly understood \citep{Guitet2018, Molino2001}. Given the variety
of species response to disturbance and the huge diversity of tropical
forests, biodiversity trajectories following disturbance might be very
diverse \citep{Lindenmayer2012, Garcia_florez2017}. We assumed that the
rate, endpoint and duration of the trajectories depend on the intensity
and frequency of disturbance: we then aimed to highlight the link
between biodiversity and disturbance and thus anticipate community
post-disturbance trajectories \citep{Chazdon2003a, Willig2018}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events \citep{Connell1978}. The theory and
states that community diversity peaks at intermediate level of
disturbance. The fluctuations of environmental conditions after
disturbance are assumed to prevent competitive exclusion in fostering
both competitively superior species and fast colonizers
\citep{Shea2004, Pulsford2016}. In tropical forests, however,
observations of the IDH are often blurry compared to theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. The IDH may result from several
processes and in the highly diverse tree communities its pattern might
differ from those recognized otherwise. Most of the studies rely on a
single diversity metric, often the taxonomic richness
\citep{Molino2001}, which is limited to grasp all facets of the
diversity-disturbance relationship. In this study we aimed as well to
clarify the scope of the IDH and discuss its implication for the fate of
tropical forests \citep{Sheil2003, Shea2004}.

The analyse of community response to disturbance requires a set of
metrics, to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
\citep{Lavorel2002, Bellwood2006} and assesses the different species
fostered by the disturbance. Diversity metrics encompassing taxonomic
richness and evenness would assess the changes community abundance
distribution. A functional approach accounting for species biological
attributes and ecological strategy, would directly link community
diversity, composition and redundancy to ecosystem functioning and
environmental constraints \citep{Violle2007b, Baraloto2012a}. In that
respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, (i) from post-disturbance trajectories we examined
community taxonomic and functional response, (ii) then we discussed the
scope of the IDH in taxonomic and functional facets and in time after
disturbance, and (iii) we analyzed community resilience and time to
recovery.

@@

\subsection{Decoupled taxonomic and functional composition
trajectories}\label{decoupled-taxonomic-and-functional-composition-trajectories}

@@

\subsection{The scope of the intermediate disturbance
hypothesis}\label{the-scope-of-the-intermediate-disturbance-hypothesis}

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
for which the taxonomic richness was maximized and the taxonomic
evenness remained resilient. The disturbance intensity determined the
balance in the community between pre-disturbance surviving trees, and
trees recruited afterward. Below a 20\%-25\% AGB loss, the trees
surviving after disturbance remained numerous enough to maintain the
high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pionners, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they balanced the usual
hyper-dominance of tropical forests, hence temporarily increasing the
taxonomic evenness \citep{Baraloto2012a}. Above the intensity threshold,
the taxonomic richness of surviving trees was too low to be offset by
the recruitment of pioneers. In the Guiana shield indeed, the pool of
true pioneers specifically recruited after disturbance is restricted to
a few common genera (e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)
\citep{Guitet2018}. The taxonomic richness then decreased following
intense disturbance, all the more so that the disturbance intensity was
high \citep{Molino2001}. At the same time, pioneers became persistently
dominant and prevented a return towards the initial evenness and the
recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity
besides highlighted an intermediate time after disturbance, around 15-20
years, for which the taxonomic evenness was maximized. As already
observed in the Guiana shield \citep{Baraloto2012a} and in Borenan
tropical forests \citep{Cannon1998}, taxonomic evenness followed
hump-shaped trajectories with a peak when the recruitment was balanced
between pioneers and late-successional species. This intermediate time
corresponded to the match between the generation time of pionneers,
first to dominate the post-disturbance recruitment, and the time for
competitive exclusion to emerge, which leads to the dominance of
late-successional species.
